Score-based tests of measurement invariance: Use in practice

### 2013-09-13 frontiers/frontiers.Rnw
update cat2bb, paragraph, change fplot option, change sim1 plots and move sim2 plots
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@258 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@

## Unrestricted model
full <- cfa('f1=~gq6_1 + gq6_2 + gq6_3 + gq6_4 + gq6_5',
             data=yg, group="agegroup", meanstructure=TRUE)
@
Finally, we can get the results of a likelihood ratio test via the \verb+anova()+ 
function, which implies that the GQ-6 violates measurement invariance.
<<LRT>>=
## LRT
anova(full, reduced)
@
Because our sample size is large, this likelihood ratio test is sensitive to 
small measurement invariance violations \cite{Ben80}.  Thus, the above test may 
be significant solely due to the large sample size.  Additionally, the above test 
is sensitive to small measurement invariance violations that are not likely to be 
of interest to researchers.  For example, imagine that the 15-year-olds' 
parameters are slightly different than the other age groups.  
The 15-year-olds are in the middle of the age groups, and there 
is not likely to be any theoretical justification for 15-year-olds 
differing from every other age group.  We need test statistics that 
are sensitive to measurement invariance violations that are monotonic 
with age.  The score-based statistics described previously possess this property.

%% TODO Illustrate catL2BB statistic?

@@

<<Get Estimate>>=
dm <- sctest(reduced, as.numeric(yg$agegroup), parm=1:4, stat="DM")
cvm <- sctest(reduced, as.numeric(yg$agegroup), parm=1:4, stat="CvM")
maxlm <- sctest(reduced, as.numeric(yg$agegroup), parm=1:4, stat="maxLM")

round(c(dm$p.value, cvm$p.value, maxlm$p.value), 2)
@
In the code above, the \verb+parm=1:4+ argument means that 
we wish to test model parameter numbers 1 through 4 (which are the factor loadings), 
which are supplied by lavaan.  The ordering of parameters can be obtained via the 
command \verb+coef(reduced)+.  We see that two of the three $p$-values output at 
the end of the code are larger than that associated with the LRT (with the CvM statistic being non-significant).

@@

<<ordtest,eval=FALSE>>=
wdmo <- sctest(reduced, as.numeric(yg$agegroup), parm=1:4, stat="WDMo")
maxlmo <- sctest(reduced, as.numeric(yg$agegroup), parm=1:4, stat="maxLMo")
@ 
<<pvals>>=
round(c(wdmo$p.value, maxlmo$p.value), 2)
@

@@

%Or alternatively, we could use the sctest.lavaan function, but assign ordinal statistics.
\begin{figure}[H]
\caption{Fluctuation processes for the WDM statistic(left panel) and the maxLM statistic(right panel)}
\label{fig:ordres}
\setkeys{Gin}{width=\textwidth}
<<ordres, fig=TRUE, height=3, width=6, echo=FALSE>>=
par(mfrow = c(1, 2), cex = 0.65)
WDMo<-sctest(reduced, as.numeric(yg$agegroup),parm=1:4, stat="WDMo", fplot=TRUE)
#WDMo 
maxLMo<-sctest(reduced, as.numeric(yg$agegroup),parm=1:4, stat="maxLMo", fnl=opval1, fplot=TRUE)
#maxLMo 
@
\end{figure}

@@

%% TODO Describe graphs
%% Is there some argument for fplot=TRUE so user can change  axis-name and critical value line
Plots representing the statistics' fluctuations across levels of age group are displayed in Figure~\ref{fig:ordres}. The left panel displays the process associated with WDM. The right panel displays the process associated with maxLM. In both cases, the test statistics in the sequence assess a split of the observations up to age group $i$ vs. greater than $i$, and the null hypothesis is rejected if the maximum of the statistics is larger than its 5\% critical value(visualized by the horizontal red line). The final group is not displayed because it encompass all observations and always equals to zero. It is observed that both statistics generally increase with age. 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\begin{figure}[H]
\caption{Simulated power curves for the ordered 
      $\max\mathit{LM}$ tests, the double-max test and the unordered $\max\mathit{LM}$ across three
      levels of the ordinal variable $m$, and measurement invariance 
      violations of 0--4 standard errors (scaled by $\sqrt{n}$). The violation parameter is Loading1.}
\label{fig:sim11res}
\setkeys{Gin}{width=\textwidth}
<<sim11res, fig=TRUE, height=7, width=7>>=
sim1.tmp <- sim1[sim1$test %in% c("ordmax","ordwmax","catdiff")&sim1$parms %in% c("loading"),]
sim1.tmp$test <- factor(sim1.tmp$test)
trellis.par.set(theme = canonical.theme(color = FALSE))
mykey <- simpleKey(levels(sim1.tmp$test), points = TRUE, lines = TRUE)
print(xyplot(power ~ diff | nlevels+pars, group = ~ test, data = sim1.tmp, 
             type = "b", xlab="Violation Magnitude", ylab="Power", key=mykey))
@

@@

\begin{figure}[H]
\caption{Simulated power curves for the ordered 
      $\max\mathit{LM}$ tests, the ordered double-max test the unordered $\max\mathit{LM}$ across three
      levels of the ordinal variable $m$, and measurement invariance
      violations of 0--4 standard errors (scaled by $\sqrt{n}$). The violation parameter is Covariance}
\label{fig:sim12res}
\setkeys{Gin}{width=\textwidth}
<<sim12res, fig=TRUE, height=7, width=7>>=
sim1.tmp <- sim1[sim1$test %in% c("ordmax","ordwmax","catdiff")&sim1$parms %in% c("var"),]
sim1.tmp$test <- factor(sim1.tmp$test)
trellis.par.set(theme = canonical.theme(color = FALSE))
mykey <- simpleKey(levels(sim1.tmp$test), points = TRUE, lines = TRUE)
print(xyplot(power ~ diff | nlevels+pars, group = ~ test, data = sim1.tmp, 
             type = "b", xlab="Violation Magnitude", ylab="Power", key=mykey))
@

@@

\begin{figure}[H]
\caption{Simulated power curves for the ordered 
      $\max\mathit{LM}$ tests, the ordered double-max test the unordered $\max\mathit{LM}$ across three
      levels of the ordinal variable $m$, and measurement invariance
      violations of 0--4 standard errors (scaled by $\sqrt{n}$). The violation parameter is Error1}
\label{fig:sim13res}
\setkeys{Gin}{width=\textwidth}
<<sim13res, fig=TRUE, height=7, width=7>>=
sim1.tmp <- sim1[sim1$test %in% c("ordmax","ordwmax","catdiff")&sim1$parms %in% c("error"),]
sim1.tmp$test <- factor(sim1.tmp$test)
trellis.par.set(theme = canonical.theme(color = FALSE))
mykey <- simpleKey(levels(sim1.tmp$test), points = TRUE, lines = TRUE)
print(xyplot(power ~ diff | nlevels+pars, group = ~ test, data = sim1.tmp, 
             type = "b", xlab="Violation Magnitude", ylab="Power", key=mykey))
@

@@

if(file.exists("sim2.rda")) {
  load("sim2.rda")
} else {
  sim2 <- simulation(sim="sim2", nobs=480, parms=c("extra","extra+loading","extra+var","extra+error"))
  sim2$nlevels <- factor(sim2$nlevels)
  levels(sim2$nlevels) <- paste("m=",levels(sim2$nlevels),sep="")
  sim2$nobs <- factor(sim2$nobs)
  levels(sim2$nobs) <- paste("n=",levels(sim2$nobs),sep="")
  levels(sim2$test) <- c("ordmax","ordwmax","catdiff")
  save(sim2, file="sim2.rda")
}
sim2$test <- factor(as.character(sim2$test),
  levels = c("ordmax","ordwmax","catdiff"),
  labels = c("ordmax","ordwmax","catdiff"))
  levels(sim2$pars) <- c("Loading1","Covariance","Error1","All Loadings","All Errors") 
@

\begin{figure}[H]
\caption{Simulated power curves for the ordered 
      $\max\mathit{LM}$ tests, the double-max test and the unordered $\max\mathit{LM}$ across three
      levels of the ordinal variable $m$, and measurement invariance 
      violations of 0--4 standard errors (scaled by $\sqrt{n}$). The violation parameter is the lacking loading.}
\label{fig:sim21res}
\setkeys{Gin}{width=\textwidth}
<<sim21res, fig=TRUE, height=7, width=7>>=
sim2.tmp <- sim2[sim2$test %in% c("ordmax","ordwmax","catdiff")&sim2$parms %in% c("extra"),]
sim2.tmp$test <- factor(sim2.tmp$test)
trellis.par.set(theme = canonical.theme(color = FALSE))
mykey <- simpleKey(levels(sim2.tmp$test), points = TRUE, lines = TRUE)
print(xyplot(power ~ diff | nlevels+pars, group = ~ test, data = sim2.tmp, 
             type = "b", xlab="Violation Magnitude", ylab="Power", key=mykey))
@

@@

\begin{figure}[H]
\caption{Simulated power curves for the ordered 
      $\max\mathit{LM}$ tests, the double-max test and the unordered $\max\mathit{LM}$ across three
      levels of the ordinal variable $m$, and measurement invariance 
      violations of 0--4 standard errors (scaled by $\sqrt{n}$). The violation parameter is the lacking loading.}
\label{fig:sim22res}
\setkeys{Gin}{width=\textwidth}
<<sim22res, fig=TRUE, height=7, width=7>>=
sim2.tmp <- sim2[sim2$test %in% c("ordmax","ordwmax","catdiff")&sim2$parms %in% c("extra+loading"),]
sim2.tmp$test <- factor(sim2.tmp$test)
trellis.par.set(theme = canonical.theme(color = FALSE))
mykey <- simpleKey(levels(sim2.tmp$test), points = TRUE, lines = TRUE)
print(xyplot(power ~ diff | nlevels+pars, group = ~ test, data = sim2.tmp, 
             type = "b", xlab="Violation Magnitude", ylab="Power", key=mykey))
@

@@

\end{figure}
\section{General Discussion}

>>>>

%% TODO Illustrate catL2BB statistic

If $V$ is only nominal/categorical,
there is not even a partial ordering, i.e., measurement invariance tests should neither
exploit the ordering of $V$'s levels nor of the observations within the level.
In this situation, it is possible to obtain a test statistic by first summing scores within
each of the $m$ levels of the auxiliary variable, then ``summing the sums'' to obtain a test
statistic \cite{HjoKon02}.  This test statistic can be formally written as
\begin{equation}
    \label{eq:lmuo}
     \mathit{LM}_\mathit{uo} = \sum_{\ell = 1, \dots, m} \sum_{j = 1, \dots, k}
       \left( {\bm B}(\hat {\bm \theta})_{i_\ell j} - {\bm B}(\hat {\bm \theta})_{i_{\ell - 1}j} \right)^2,
\end{equation}
where the first subscript on the two terms in parentheses are $i_\ell$
and $i_{\ell-1}$, respectively (and where we take $i_0 = 0$, so that
the cumulative score is ${\bm B}(0, \hat {\bm \theta}) = \bm{0}$).
Again, tests of subsets of model parameters can be obtained by 
taking the inner sum over only the $k^* < k$ parameters of interest.
This test statistic discards the ordinal nature of the auxiliary
variable, essentially employing the alternative hypothesis
from~\eqref{eq:h2*}.  A similar issue is observed in testing for
measurement 
invariance via multiple groups models and likelihood ratio tests (or,
equivalently, via Wald tests or Lagrange multiplier tests): we
can allow $\bm{\theta}$ to be unique at each level of the ordinal
variable, but the ordinality of the auxiliary variable is lost.

@@

## Unrestricted model
full <- cfa('f1=~gq6_1 + gq6_2 + gq6_3 + gq6_4 + gq6_5',
             data=yg, group="agegroup", meanstructure=TRUE)
@
Finally, we can get the results of a likelihood ratio test via the \verb+anova()+ 
function, which implies that the GQ-6 violates measurement invariance.
<<LRT>>=
## LRT
anova(full, reduced)
@

Another similar statistics is unordered $\mathit{LM}_\mathit{uo}$ as shown in Equ. We can obtain this statistic by using the following function. 
<<Categorical>>=
## unordered LM
catdiff <- sctest(reduced, as.numeric(yg$agegroup), parm=1:4, stat="range")
round(c(catdiff$p.value), 2)
@
Because our sample size is large, this likelihood ratio test is sensitive to 
small measurement invariance violations \cite{Ben80}.  Thus, the above test may 
be significant solely due to the large sample size.  Additionally, the above test 
is sensitive to small measurement invariance violations that are not likely to be 
of interest to researchers.  For example, imagine that the 15-year-olds' 
parameters are slightly different than the other age groups.  
The 15-year-olds are in the middle of the age groups, and there 
is not likely to be any theoretical justification for 15-year-olds 
differing from every other age group.  We need test statistics that 
are sensitive to measurement invariance violations that are monotonic 
with age.  The score-based statistics described previously possess this property.

@@

<<Get Estimate>>=
dm <- sctest(reduced, as.numeric(yg$agegroup), parm=1:4, stat="DM")
cvm <- sctest(reduced, as.numeric(yg$agegroup), parm=1:4, stat="CvM")
maxlm <- sctest(reduced, as.numeric(yg$agegroup), parm=1:4, stat="maxLM")
round(c(dm$p.value, cvm$p.value, maxlm$p.value), 2)
@
In the code above, the \verb+parm=1:4+ argument means that 
we wish to test model parameter numbers 1 through 4 (which are the factor loadings), 
which are supplied by lavaan.  The ordering of parameters can be obtained via the 
command \verb+coef(reduced)+.  We see that two of the three $p$-values output at 
the end of the code are larger than that associated with the LRT (with the CvM statistic being non-significant).

@@

<<ordtest,eval=FALSE>>=
wdmo <- sctest(reduced, as.numeric(yg$agegroup), parm=1:4, stat="WDMo")
maxlmo <- sctest(reduced, as.numeric(yg$agegroup), parm=1:4, stat="maxLMo")

@ 
<<pvals>>=
round(c(wdmo$p.value, maxlmo$p.value), 2)
@

@@

%Or alternatively, we could use the sctest.lavaan function, but assign ordinal statistics.
\begin{figure}[H]
\caption{Fluctuation processes for the WDM statistic(left panel) and the maxLM statistic(right panel)}
\label{fig:ordres}
\setkeys{Gin}{width=\textwidth}
<<ordres, fig=TRUE, height=3, width=6, echo=FALSE>>=
par(mfrow = c(1, 2), cex = 0.65)
WDMo<-sctest(reduced, as.numeric(yg$agegroup),parm=1:4, stat="WDMo", fplot=TRUE,ylim=c(0,3.5),xlab="Age group",ylab = "Weighted max statistics")

#WDMo 
maxLMo<-sctest(reduced,as.numeric(yg$agegroup),parm=1:4, stat="maxLMo", fnl=opval1, fplot=TRUE,ylim=c(0,14),xlab="Age group",ylab = "LM statistics")
#maxLMo 
@
\end{figure}

\seeme{can't change age group axis, since order.by argument needs agegroup as numeric}

@@

%% TODO Describe graphs
%% Is there some argument for fplot=TRUE so user can change  axis-name and critical value line
Plots representing the statistics' fluctuations across levels of age group are displayed in Figure~\ref{fig:ordres}. The left panel displays the process associated with WDM. The right panel displays the process associated with maxLM. In both cases, the test statistics in the sequence assess a split of the observations up to age group $i$ vs. greater than $i$, and the null hypothesis is rejected if the maximum of the statistics is larger than its 5\% critical value(visualized by the horizontal red line). The final group is not displayed because it encompass all observations and always equals to zero. The age group 1:5 in the x-axis corresponds to 10-11, 12-13, 14,15,16 respectively. It is observed that both statistics generally increase with age. 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\begin{figure}[H]
\caption{Simulated power curves for the ordered 
      $\max\mathit{LM}$ tests, the double-max test and the unordered $\max\mathit{LM}$ across three
      levels of the ordinal variable $m$, and measurement invariance 
      violations of 0--4 standard errors (scaled by $\sqrt{n}$). The violation parameter is Loading1.}
\label{fig:sim11res}
\setkeys{Gin}{width=\textwidth}
<<sim11res, fig=TRUE, height=7, width=7>>=
sim1.tmp <- sim1[sim1$test %in% c("ordmax","ordwmax","catdiff")&sim1$parms %in% c("loading")&sim1$pars %in% c("Loading1","All Loadings","All Errors"),]
sim1.tmp$test <- factor(sim1.tmp$test)
trellis.par.set(theme = canonical.theme(color = FALSE))
mykey <- simpleKey(levels(sim1.tmp$test), points = TRUE, lines = TRUE)
print(xyplot(power ~ diff | nlevels+pars, group = ~ test, data = sim1.tmp, 
             type = "b", xlab="Violation Magnitude", ylab="Power", key=mykey))
@

@@

\begin{figure}[H]
\caption{Simulated power curves for the ordered 
      $\max\mathit{LM}$ tests, the ordered double-max test the unordered $\max\mathit{LM}$ across three
      levels of the ordinal variable $m$, and measurement invariance
      violations of 0--4 standard errors (scaled by $\sqrt{n}$). The violation parameter is Covariance}
\label{fig:sim12res}
\setkeys{Gin}{width=\textwidth}
<<sim12res, fig=TRUE, height=7, width=7>>=
sim1.tmp <- sim1[sim1$test %in% c("ordmax","ordwmax","catdiff")&sim1$parms %in% c("var") &sim1$pars %in% c("All Loadings", "Covariance","All Errors") ,]
sim1.tmp$test <- factor(sim1.tmp$test)
trellis.par.set(theme = canonical.theme(color = FALSE))
mykey <- simpleKey(levels(sim1.tmp$test), points = TRUE, lines = TRUE)
print(xyplot(power ~ diff | nlevels+pars, group = ~ test, data = sim1.tmp, 
             type = "b", xlab="Violation Magnitude", ylab="Power", key=mykey))
@

@@

\begin{figure}[H]
\caption{Simulated power curves for the ordered 
      $\max\mathit{LM}$ tests, the ordered double-max test the unordered $\max\mathit{LM}$ across three
      levels of the ordinal variable $m$, and measurement invariance
      violations of 0--4 standard errors (scaled by $\sqrt{n}$). The violation parameter is Error1}
\label{fig:sim13res}
\setkeys{Gin}{width=\textwidth}
<<sim13res, fig=TRUE, height=7, width=7>>=
sim1.tmp <- sim1[sim1$test %in% c("ordmax","ordwmax","catdiff")&sim1$parms %in% c("error") &sim1$pars %in% c("Error1","All Errors","All Loadings"),]
sim1.tmp$test <- factor(sim1.tmp$test)
trellis.par.set(theme = canonical.theme(color = FALSE))
mykey <- simpleKey(levels(sim1.tmp$test), points = TRUE, lines = TRUE)
print(xyplot(power ~ diff | nlevels+pars, group = ~ test, data = sim1.tmp, 
             type = "b", xlab="Violation Magnitude", ylab="Power", key=mykey))
@

@@

if(file.exists("sim2.rda")) {
  load("sim2.rda")
} else {
  sim2 <- simulation(sim="sim2", nobs=480, parms=c("extra","extra+loading","extra+var","extra+error"))
  sim2$nlevels <- factor(sim2$nlevels)
  levels(sim2$nlevels) <- paste("m=",levels(sim2$nlevels),sep="")
  sim2$nobs <- factor(sim2$nobs)
  levels(sim2$nobs) <- paste("n=",levels(sim2$nobs),sep="")
  levels(sim2$test) <- c("ordmax","ordwmax","catdiff")
  save(sim2, file="sim2.rda")
}
sim2$test <- factor(as.character(sim2$test),
  levels = c("ordmax","ordwmax","catdiff"),
  labels = c("ordmax","ordwmax","catdiff"))
  levels(sim2$pars) <- c("Loading1","Covariance","Error1","All Loadings","All Errors") 
  levels(sim2$nlevels) <- paste("m=",levels(sim2$nlevels),sep="")
@

\begin{figure}[H]
\caption{Simulated power curves for the ordered 
      $\max\mathit{LM}$ tests, the double-max test and the unordered $\max\mathit{LM}$ across three
      levels of the ordinal variable $m$, and measurement invariance 
      violations of 0--4 standard errors (scaled by $\sqrt{n}$). The violation parameter is the lacking loading.}
\label{fig:sim21res}
\setkeys{Gin}{width=\textwidth}
<<sim21res, fig=TRUE, height=7, width=7>>=
sim2.tmp <- sim2[sim2$test %in% c("ordmax","ordwmax","catdiff")&sim2$parms %in% c("extra")&sim2$pars %in% c("Loading1","Error1","Covariance"),]
sim2.tmp$test <- factor(sim2.tmp$test)
trellis.par.set(theme = canonical.theme(color = FALSE))
mykey <- simpleKey(levels(sim2.tmp$test), points = TRUE, lines = TRUE)
print(xyplot(power ~ diff | nlevels+pars, group = ~ test, data = sim2.tmp, 
             type = "b", xlab="Violation Magnitude", ylab="Power", key=mykey))
@

@@

\section{General Discussion}

\section{Appendix}
\begin{figure}[H]
\caption{Simulated power curves for the ordered 
      $\max\mathit{LM}$ tests, the double-max test and the unordered $\max\mathit{LM}$ across three
      levels of the ordinal variable $m$, and measurement invariance 
      violations of 0--4 standard errors (scaled by $\sqrt{n}$). The violation parameter is the lacking loading.}
\label{fig:sim22res}
\setkeys{Gin}{width=\textwidth}
<<sim22res, fig=TRUE, height=7, width=7>>=
sim2.tmp <- sim2[sim2$test %in% c("ordmax","ordwmax","catdiff")&sim2$parms %in% c("extra+loading"),]
sim2.tmp$test <- factor(sim2.tmp$test)
trellis.par.set(theme = canonical.theme(color = FALSE))
mykey <- simpleKey(levels(sim2.tmp$test), points = TRUE, lines = TRUE)
print(xyplot(power ~ diff | nlevels+pars, group = ~ test, data = sim2.tmp, 
             type = "b", xlab="Violation Magnitude", ylab="Power", key=mykey))
@

@@
